---
title: "Complete Case Study: From Gene Sets to Discovery"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Complete Case Study: From Gene Sets to Discovery}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  eval = TRUE,
  collapse = TRUE,
  comment = "#>"
)
```

## Overview

This case study demonstrates a complete GPSA workflow, from simple gene set queries to complex multi-channel event discovery. We'll use a realistic biological scenario to showcase the full power of GPSA.

**Scenario**: You have identified a set of genes differentially expressed in your experiment and want to find perturbations in public databases that produce similar or related effects.

## Tutorial Roadmap

GPSA supports increasingly sophisticated query types. Here's how they build on each other:

```
Single Gene Set Query (Get Started)
         |
         v
Up/Down Bidirectional Query -----> Find similar/opposite perturbations
         |
         v
Multi-channel Event Definition --> Complex phenotype queries
         |
         v
Batch Queries + Meta-analysis ---> Compare multiple signatures
         |
         v
Full Vector Search --------------> Complete expression profile queries
```

## Detailed Tutorials

Each tutorial covers a specific capability in depth:

### 1. Up/Down Bidirectional Query

**[Tutorial: Up/Down Query](tutorial_updown.html)**

When you have a reference perturbation signature (with both up- and down-regulated genes), you can find other perturbations that produce **similar** or **opposite** effects.

Key concepts covered:
- Fixed gene selection (`n_each = 500`)
- Adaptive gene selection based on energy threshold
- Phenocopy vs Discovery scoring modes
- Channel-level score interpretation

### 2. Multi-channel Event Definition

**[Tutorial: Viral Mimicry](tutorial_viral_mimicry.html)**

Define complex biological events by combining multiple gene sets with different expected directions. Example: "viral mimicry" state requires both interferon activation AND cell cycle suppression.

Key concepts covered:
- Multi-channel query construction
- Gated filtering (all channels must satisfy direction)
- Dual-mode scoring (Phenocopy vs Discovery)
- Biological interpretation of multi-channel hits

### 3. Batch Queries and Meta-analysis

**[Tutorial: Batch Queries](tutorial_batch.html)**

When you have multiple related signatures (e.g., replicates or related conditions), run them together for efficiency and perform meta-analysis.

Key concepts covered:
- `gpsa_batch_search()` for multiple queries
- Query-to-query correlation analysis
- Stouffer's method for meta-analysis
- Identifying consensus hits across queries

### 4. Full Vector Search

**[Tutorial: Vector Search](tutorial_vector_search.html)**

Use complete gene expression profiles (e.g., full logFC vectors) as queries, leveraging the entire transcriptomic information rather than just top/bottom genes.

Key concepts covered:
- `gpsa_search_vector()` with rankZ transformation
- Adaptive sparsification (energy-based gene selection)
- Compile information interpretation
- Comparison with Up/Down approach

## Quick Reference: Choosing the Right Query Type

| Scenario | Recommended Approach | Tutorial |
|----------|---------------------|----------|
| Query with a pathway gene set | Single gene set query | [Get Started](GPSA.html) |
| Find perturbations similar to a reference | Up/Down query | [Up/Down](tutorial_updown.html) |
| Complex phenotype (multiple pathways) | Multi-channel query | [Viral Mimicry](tutorial_viral_mimicry.html) |
| Multiple related signatures | Batch + meta-analysis | [Batch](tutorial_batch.html) |
| Complete expression profile | Vector search | [Vector Search](tutorial_vector_search.html) |

## Performance Tips

1. **Use memory mode** for interactive analysis:
   ```r
   db <- gpsa_load_db(preload_Z = TRUE)
   ```

2. **Batch related queries** to share I/O overhead:
   ```r
   bat <- gpsa_batch_search(db, list(Q1 = q1, Q2 = q2, Q3 = q3))
   ```

3. **Use adaptive gene selection** to avoid arbitrary cutoffs:
   ```r
   query <- gpsa_make_updown_query_adaptive(gl, eta = 0.90)
   ```

4. **Apply dual-mode scoring** to distinguish phenocopy from discovery:
   ```r
   out <- gpsa_postprocess_modes(out, p_mode = "norm")
   head(out$res_similarity)  # Phenocopy hits
   head(out$res_discovery)   # Discovery hits
   ```

## API Quick Reference

### Core Functions

| Function | Description |
|----------|-------------|
| `gpsa_load_db()` | Load HDF5 database |
| `gpsa_search()` | Gene-set query |
| `gpsa_search_vector()` | Full geneList query |
| `gpsa_batch_search()` | Multiple queries at once |

### Query Helpers

| Function | Description |
|----------|-------------|
| `gpsa_make_updown_query()` | Create Up/Down query (fixed n) |
| `gpsa_make_updown_query_adaptive()` | Adaptive gene selection |
| `gpsa_get_signature()` | Extract signature vector |

### Post-processing

| Function | Description |
|----------|-------------|
| `gpsa_postprocess_modes()` | Dual-mode scoring |
| `gpsa_attach_channel_scores()` | Add per-channel columns |
| `gpsa_meta_stouffer()` | Meta-analysis across queries |

### Visualization

| Function | Description |
|----------|-------------|
| `gpsa_plot_gsea()` | GSEA running enrichment plot |
| `gpsa_gsea_running()` | Compute running scores |

## Session Info

```{r session}
sessionInfo()
```
