---
title: "Get Started with GPSA"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Get Started with GPSA}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(eval = FALSE)
```

## What is GPSA?
GPSA (Gene-set Perturbation Signature Analysis) provides **fast similarity search** over large-scale perturbation signature databases using cMAP/GSEA-style rankZ semantics.

**Key Features:**

- Single gene-set queries
- Up/Down bidirectional queries (traditional cMAP-style)
- Multi-channel event definitions (e.g., viral mimicry)
- Dual-mode scoring: Phenocopy (strict) vs Discovery (one-sided)
- Batch queries with meta-analysis
- Full geneList vector search with adaptive sparsification
- HDF5-based storage for memory efficiency

## Installation

```{r install}
# Set mirrors (China users)
options("repos" = c(CRAN = "https://mirrors.tuna.tsinghua.edu.cn/CRAN/"))
options(BioC_mirror = "https://mirrors.westlake.edu.cn/bioconductor")

# Install dependencies
if (!requireNamespace("BiocManager", quietly = TRUE)) install.packages("BiocManager")
BiocManager::install("rhdf5")
install.packages("Matrix")

# Install GPSA from local source
install.packages("/path/to/GPSA", repos = NULL, type = "source")

# Or use devtools
# devtools::install_local("/path/to/GPSA")
```

## First-time Setup: HDF5 Database

GPSA requires a pre-built HDF5 database containing perturbation signatures. Download the database file and set its path:

```{r setup-h5}
library(GPSA)

# Method 1: R option (recommended, persists for session)
options(GPSA.h5 = "/path/to/gpsa_db.h5")

# Method 2: Environment variable
# Sys.setenv(GPSA_H5 = "/path/to/gpsa_db.h5")

# Method 3: Specify h5file parameter in each call
# db <- gpsa_load_db(h5file = "/path/to/gpsa_db.h5")
```

## Loading the Database

GPSA offers two loading modes:

### Memory Mode (Recommended for interactive analysis)

```{r memory-mode}
# Preload Z matrix into RAM (~1GB, ~9s load time)
# Subsequent queries: ~0.15s each
db <- gpsa_load_db(preload_Z = TRUE, verbose = TRUE)
```

### Disk Mode (For limited memory)

```{r disk-mode}
# No preload, read from disk on each query
# Load time: ~0.04s, Query time: ~7s each
db <- gpsa_load_db(preload_Z = FALSE, verbose = TRUE)
```

## Quick Example: Single Gene Set Query

```{r single-query}
# Define a query gene set
query <- list(
  MyGenes = list(
    genes = c("TP53", "BRCA1", "EGFR", "MYC", "KRAS"),  # Your genes
    sign = +1,    # +1 for up-regulation, -1 for down-regulation
    weight = 1    # Optional weight
  )
)

# Search
out <- gpsa_search(db, query, return_components = FALSE)

# View results
head(out$res[, c("signature_id", "pbgene", "CellLineName", "Score", "tau", "FDR_rank")])
```

## Quick Example: Up/Down Query

Find perturbations similar to a reference signature:

```{r updown-query}
# Get a reference signature's logFC
gl <- gpsa_get_signature(db, "D21455", dataset = "logFC")

# Create Up/Down query (top 300 + bottom 300 genes)
query_ud <- gpsa_make_updown_query(gl, n_each = 300)

# Search
out <- gpsa_search(db, query_ud, return_components = TRUE)

# Most similar perturbations
head(out$res)

# Most opposite perturbations
tail(out$res)
```

## Performance Summary

| Mode | Query Type | Time (avg) |
|------|------------|------------|
| Memory | Single gene set | 0.14s |
| Memory | Up/Down 2-channel | 0.20s |
| Memory | Multi 4-channel | 0.25s |
| Memory | Batch 5 queries | 0.47s |
| Disk | Single gene set | 7.2s |

**Recommendation**: Use `preload_Z = TRUE` for interactive analysis.

## Next Steps

For complete examples including:

- **Estrogen pathway query** - Single gene set with GSEA validation
- **Viral mimicry event definition** - Multi-channel gated search
- **Dual-mode scoring** - Phenocopy vs Discovery
- **Batch queries and meta-analysis**
- **Full vector search with adaptive sparsification**

See the **[Complete Case Study](case_study.html)** vignette.

## API Reference

### Core Functions

| Function | Description |
|----------|-------------|
| `gpsa_load_db()` | Load HDF5 database |
| `gpsa_search()` | Gene-set query |
| `gpsa_search_vector()` | Full geneList query |
| `gpsa_batch_search()` | Multiple queries at once |
| `gpsa_build_db()` | Build new database |

### Query Helpers

| Function | Description |
|----------|-------------|
| `gpsa_make_updown_query()` | Create Up/Down query |
| `gpsa_make_updown_query_adaptive()` | Adaptive gene selection (energy-based) |
| `gpsa_get_signature()` | Extract signature vector |

### Post-processing

| Function | Description |
|----------|-------------|
| `gpsa_postprocess_modes()` | Dual-mode scoring (Phenocopy/Discovery) |
| `gpsa_attach_channel_scores()` | Add per-channel columns |
| `gpsa_meta_stouffer()` | Meta-analysis across queries |

### Visualization

| Function | Description |
|----------|-------------|
| `gpsa_plot_gsea()` | GSEA running enrichment plot |
| `gpsa_gsea_running()` | Compute running scores |

## Session Info

```{r session}
sessionInfo()
```

