---
title: "Tutorial: Batch Queries and Meta-analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Tutorial: Batch Queries and Meta-analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
# Resolve HDF5 database path (for vignette rendering)
resolve_h5_path <- function() {
  h5 <- Sys.getenv("GPSA_H5", unset = "")
  if (h5 != "" && file.exists(h5)) return(h5)
  
  h5 <- getOption("GPSA.h5", default = "")
  if (h5 != "" && file.exists(h5)) return(h5)
  
  candidates <- c(
    "../output/gpsa_rankZ_logFC_optimized.h5",
    "output/gpsa_rankZ_logFC_optimized.h5",
    "~/gpsa_db.h5"
  )
  for (h5 in candidates) {
    if (file.exists(h5)) return(normalizePath(h5))
  }
  return("")
}

h5_path <- resolve_h5_path()
has_db <- (h5_path != "")

knitr::opts_chunk$set(
  eval = has_db,
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)

if (has_db) {
  options(GPSA.h5 = h5_path)
}
```

## Overview

When you have multiple related signatures (e.g., biological replicates, time points, or related conditions), running them individually is inefficient. GPSA's batch query feature:

- Runs multiple queries in a single matrix operation
- Computes query-to-query correlations
- Enables meta-analysis to find consensus hits

## Setup

```{r load}
library(GPSA)

# Load database
db <- gpsa_load_db(preload_Z = TRUE, verbose = TRUE)
```

## Preparing Multiple Queries

### Select Related Signatures

```{r select-signatures}
# Choose multiple related signatures
# Let's use the first 3 signatures from the database for this demo
sig_ids <- db$sigs[1:3]

message(sprintf("Using %d signatures for batch query:", length(sig_ids)))
print(sig_ids)
```

### Build Query List

```{r build-queries}
# Create Up/Down query for each signature
queries <- list()
for (sid in sig_ids) {
  gl <- gpsa_get_signature(db, sid, dataset = "logFC")
  queries[[sid]] <- gpsa_make_updown_query(gl, n_each = 300)
}

# Preview query structure
message("Query names:")
print(names(queries))
```

## Running Batch Search

```{r batch-search}
# Run all queries at once
bat <- gpsa_batch_search(db, queries, compute_tau = TRUE)

# View timing information
message("=== Batch Query Timing ===")
print(bat$timing)
```

### Batch Results Structure

```{r batch-structure}
# The result contains:
# - Score: matrix of scores (signatures x queries)
# - tau: matrix of tau scores
# - z_ref: matrix of z-scores for meta-analysis
# - timing: execution time

message(sprintf("Score matrix dimensions: %d signatures x %d queries", 
                nrow(bat$Score), ncol(bat$Score)))
message("Query names:")
print(colnames(bat$Score))
```

### View Results for Each Query

```{r individual-results}
# Get results for a specific query
query_name <- names(queries)[1]
scores_q1 <- bat$Score[, query_name]

# Sort to get top hits for this query
top_q1 <- sort(scores_q1, decreasing = TRUE)[1:10]
message(sprintf("=== Top 10 for Query: %s ===", query_name))
print(top_q1)
```

## Query-to-Query Correlation

Understanding how similar your queries are helps interpret the results.

```{r query-correlation}
# Compute correlation between query results
cor_mat <- cor(bat$Score, use = "pairwise.complete.obs")

message("=== Query Similarity Matrix ===")
print(round(cor_mat, 3))
```

### Interpreting Correlations

| Correlation | Interpretation |
|-------------|----------------|
| > 0.7 | Highly similar queries (may be redundant) |
| 0.3 - 0.7 | Related but distinct queries |
| < 0.3 | Independent queries |
| Negative | Opposite effects |

## Meta-analysis: Stouffer's Method

When queries represent related conditions, combine their results to find consensus hits.

```{r meta-analysis}
# Stouffer's method combines z-scores across queries
meta_res <- gpsa_meta_stouffer(bat$z_ref)

# View meta-analysis results
message("=== Meta-analysis Top 20 ===")
print(head(meta_res[, c("signature_id", "z_meta", "p_meta", "FDR_meta")], 20))
```

### Understanding Meta-analysis Output

| Column | Description |
|--------|-------------|
| `z_meta` | Combined z-score (Stouffer's method) |
| `p_meta` | P-value from combined z-score |
| `FDR_meta` | False discovery rate (BH correction) |

**Interpretation:**
- High `z_meta`: Consistently high-scoring across all queries
- Low `FDR_meta`: Statistically significant consensus hit

## Finding Consensus Hits

```{r consensus}
# Define consensus: significant in meta-analysis
consensus <- meta_res[meta_res$FDR_meta < 0.05, ]
consensus <- consensus[order(consensus$z_meta, decreasing = TRUE), ]

message(sprintf("=== Consensus Hits (FDR < 0.05): %d ===", nrow(consensus)))
if (nrow(consensus) > 0) {
  print(head(consensus, 20))
}
```

### Compare Individual vs Meta Results

```{r compare-results}
# Check if a hit is consistent across queries
if (nrow(consensus) > 0) {
  candidate <- consensus$signature_id[1]
  
  # Get individual scores
  individual_scores <- bat$Score[candidate, ]
  message(sprintf("=== Individual Scores for %s ===", candidate))
  print(individual_scores)
  
  # Get individual tau scores
  individual_tau <- bat$tau[candidate, ]
  message("=== Individual Tau Scores ===")
  print(individual_tau)
}
```

## Advanced: Weighted Meta-analysis

If queries have different reliability, apply weights:

```{r weighted-meta, eval=FALSE}
# Example: weight by sample size or confidence
weights <- c(1.0, 0.8, 1.2)  # Adjust based on your criteria
names(weights) <- names(queries)

# Manual weighted Stouffer
z_weighted <- bat$z_ref %*% weights / sqrt(sum(weights^2))
```

## Batch Query with Different Query Types

You can mix different query types in a batch:

```{r mixed-batch, eval=FALSE}
# Prepare mixed queries
# Note: Each query in the list should be a valid query format
mixed_queries <- list(
  # Up/Down query from first signature
  UD_query = gpsa_make_updown_query(
    gpsa_get_signature(db, sig_ids[1], dataset = "logFC"), 
    n_each = 300
  ),
  # Single gene set query (example with TP53 pathway genes)
  TP53_query = list(
    TP53 = list(
      genes = c("TP53", "MDM2", "CDKN1A", "BAX", "BBC3", "PUMA"),
      sign = +1
    )
  )
)

# Run mixed batch
bat_mixed <- gpsa_batch_search(db, mixed_queries, compute_tau = TRUE)

message("=== Mixed Batch Query Timing ===")
print(bat_mixed$timing)
```

## Visualizing Query Relationships

```{r visualize, fig.cap="Heatmap showing correlation between query results."}
# Heatmap of query correlations (if ggplot2 available)
if (requireNamespace("ggplot2", quietly = TRUE)) {
  library(ggplot2)
  
  # Melt correlation matrix for plotting
  cor_df <- as.data.frame(as.table(cor_mat))
  names(cor_df) <- c("Query1", "Query2", "Correlation")
  
  print(
    ggplot(cor_df, aes(Query1, Query2, fill = Correlation)) +
      geom_tile() +
      scale_fill_gradient2(low = "blue", mid = "white", high = "red", 
                           midpoint = 0, limits = c(-1, 1)) +
      theme_minimal() +
      theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
      labs(title = "Query-to-Query Correlation")
  )
}
```

## Performance Considerations

### Batch Size

```{r batch-size, eval=FALSE}
# Batch queries share I/O overhead
# Running 5 queries in batch is much faster than 5 individual searches

# Timing comparison (illustrative)
# Individual: 5 x 0.20s = 1.0s
# Batch: ~0.47s (shared matrix operations)
```

### Memory Usage

```{r memory, eval=FALSE}
# Each query adds minimal overhead in batch mode
# The database Z matrix (~1GB) is loaded once and reused
```

## Best Practices

1. **Group related queries**: Batch queries that make biological sense together
2. **Check correlations first**: Highly correlated queries may not add information
3. **Use meta-analysis for consensus**: Stouffer's method finds robust hits
4. **Validate consensus hits**: They're more likely to be biologically meaningful
5. **Consider weights**: If query reliability varies, use weighted meta-analysis

## Summary

| Task | Function |
|------|----------|
| Batch search | `gpsa_batch_search(db, queries)` |
| Query correlation | `cor(bat$Score)` |
| Meta-analysis | `gpsa_meta_stouffer(bat$z_ref)` |

## Session Info

```{r session}
sessionInfo()
```
