---
title: "Tutorial: Up/Down Bidirectional Query"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Tutorial: Up/Down Bidirectional Query}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
# Resolve HDF5 database path (for vignette rendering)
resolve_h5_path <- function() {
  h5 <- Sys.getenv("GPSA_H5", unset = "")
  if (h5 != "" && file.exists(h5)) return(h5)
  
  h5 <- getOption("GPSA.h5", default = "")
  if (h5 != "" && file.exists(h5)) return(h5)
  
  candidates <- c(
    "../output/gpsa_rankZ_logFC_optimized.h5",
    "output/gpsa_rankZ_logFC_optimized.h5",
    "~/gpsa_db.h5"
  )
  for (h5 in candidates) {
    if (file.exists(h5)) return(normalizePath(h5))
  }
  return("")
}

h5_path <- resolve_h5_path()
has_db <- (h5_path != "")

knitr::opts_chunk$set(
  eval = has_db,
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)

if (has_db) {
  options(GPSA.h5 = h5_path)
}
```

## Overview

When you have a reference perturbation signature (with both up-regulated and down-regulated genes), you can search for other perturbations that produce **similar** or **opposite** transcriptomic effects. This is the classic Connectivity Map (cMAP) approach.

GPSA supports two methods for selecting query genes:
1. **Fixed selection**: Take top N and bottom N genes by logFC
2. **Adaptive selection**: Use energy threshold to determine optimal gene counts

## Setup

```{r load}
library(GPSA)

# Load database in memory mode
db <- gpsa_load_db(preload_Z = TRUE, verbose = TRUE)
```

## Method 1: Fixed Gene Selection

The classic approach uses a fixed number of genes from each tail of the expression distribution.

### Extract a Reference Signature

```{r ref-signature}
# Choose a reference signature (e.g., an ESR1 perturbation)
# First, let's find an ESR1-related signature
esr1_sigs <- db$meta[grep("ESR1", db$meta$pbgene, ignore.case = TRUE), ]
if (nrow(esr1_sigs) > 0) {
  sig_ref <- esr1_sigs$signature_id[1]
} else {
  sig_ref <- db$sigs[1]  # Fallback to first signature
}
message(sprintf("Using reference signature: %s", sig_ref))

# Get the logFC vector
gl_ref <- gpsa_get_signature(db, sig_ref, dataset = "logFC")

# Preview the distribution
summary(gl_ref)
```

### Create Up/Down Query (Default: 500 each)

```{r updown-500}
# Create query with top 500 up-regulated and bottom 500 down-regulated genes
query_ud <- gpsa_make_updown_query(gl_ref, n_each = 500)

message(sprintf("Up genes: %d", length(query_ud$Up$genes)))
message(sprintf("Down genes: %d", length(query_ud$Down$genes)))

# Preview the query structure
str(query_ud)
```

### Run the Search

```{r search-fixed}
out_ud <- gpsa_search(db, query_ud, return_components = TRUE)

# Check if reference ranks highly (should be near top)
self_rank <- gpsa_rank_in(out_ud$res, sig_ref)
message(sprintf("Reference signature self-rank: %d", self_rank))

# Top 10 most similar perturbations
message("=== Top 10 Most Similar ===")
print(head(out_ud$res[, c("signature_id", "pbgene", "CellLineName",
                          "Score", "tau", "FDR_rank")], 10))

# Bottom 10 most opposite perturbations
message("=== Top 10 Most Opposite ===")
print(tail(out_ud$res[, c("signature_id", "pbgene", "CellLineName",
                          "Score", "tau", "FDR_rank")], 10))
```

### Exclude Self to Find Similar Others

```{r similar-others}
# Find the most similar perturbation that isn't the reference itself
top_other <- out_ud$res[out_ud$res$signature_id != sig_ref, ][1, ]
message("=== Most Similar Other Perturbation ===")
print(top_other[, c("signature_id", "pbgene", "CellLineName", "Score", "tau")])
```

## Method 2: Adaptive Gene Selection

Instead of arbitrary cutoffs (500 genes), use energy-based thresholding to automatically determine how many genes capture a given fraction of the signal.

### Adaptive Selection with Energy Threshold

```{r adaptive}
# Select genes that capture 90% of the signal energy
q_adaptive <- gpsa_make_updown_query_adaptive(
  gl_ref,
  eta = 0.90,      # Capture 90% of energy
  min_n = 200,     # At least 200 genes per direction
  max_n = 2000     # At most 2000 genes per direction
)

# View selection information
message("=== Adaptive Selection Info ===")
print(q_adaptive$.info)

# The selected gene counts may differ from 500
message(sprintf("Adaptive Up genes: %d", length(q_adaptive$Up$genes)))
message(sprintf("Adaptive Down genes: %d", length(q_adaptive$Down$genes)))
```

### Run Search with Adaptive Query

```{r search-adaptive}
# Note: Use only Up and Down, not the .info metadata
out_adaptive <- gpsa_search(db, q_adaptive[c("Up", "Down")], 
                            return_components = TRUE)

# Compare self-rank with fixed approach
self_rank_ad <- gpsa_rank_in(out_adaptive$res, sig_ref)
message(sprintf("Adaptive self-rank: %d (vs fixed: %d)", self_rank_ad, self_rank))
```

## Examining Channel-level Scores

GPSA computes separate enrichment scores for the Up and Down channels. These can reveal whether a hit matches both directions or just one.

```{r channel-scores}
# Attach per-channel scores to results
res_channels <- gpsa_attach_channel_scores(
  out_ud$res, 
  out_ud$NES_raw, 
  out_ud$NES_signed
)

# View channel breakdown
message("=== Top 5 with Channel Scores ===")
print(head(res_channels[, c("signature_id", "pbgene", "Score",
                            "Up_raw", "Down_raw", 
                            "Up_signed", "Down_signed")], 5))
```

### Interpreting Channel Scores

| Score | Meaning |
|-------|---------|
| `Up_raw` | Raw NES for Up channel (unsigned) |
| `Down_raw` | Raw NES for Down channel (unsigned) |
| `Up_signed` | Signed NES for Up channel (+/- indicates direction match) |
| `Down_signed` | Signed NES for Down channel (+/- indicates direction match) |

A hit with **both** `Up_signed > 0` AND `Down_signed > 0` shows concordant regulation in both directions (true phenocopy).

## Dual-mode Scoring: Phenocopy vs Discovery

GPSA distinguishes two types of hits:

- **Phenocopy** (Score_bi): Both channels must show the expected direction (min of signed scores)
- **Discovery** (Score_uni): At least one channel shows strong signal (max of signed scores)

```{r dual-mode}
out_modes <- gpsa_postprocess_modes(
  out_ud,
  attach_channel_cols = TRUE,
  p_mode = "norm",
  filter_positive = TRUE
)

# Phenocopy hits: strict matches where both channels agree
message("=== Phenocopy Mode Top 10 (Score_bi) ===")
print(head(out_modes$res_similarity[, c("signature_id", "pbgene", 
                                         "Score_bi", "Score", "Score_uni",
                                         "Balance", "Driver")], 10))

# Discovery hits: strong in at least one channel
message("=== Discovery Mode Top 10 (Score_uni) ===")
print(head(out_modes$res_discovery[, c("signature_id", "pbgene",
                                        "Score_uni", "Score", "Score_bi",
                                        "Balance", "Driver")], 10))
```

### Interpreting Dual-mode Results

| Column | Description |
|--------|-------------|
| `Score_bi` | Phenocopy score: min(Up_signed, Down_signed) |
| `Score_uni` | Discovery score: max(Up_signed, Down_signed) |
| `Balance` | Ratio Score_bi / Score_uni (1 = perfectly balanced) |
| `Driver` | Which channel dominates ("Up", "Down", or "Balanced") |

**Use cases:**
- High `Score_bi` + High `Balance`: True phenocopy, both directions match
- High `Score_uni` + Low `Balance`: One-sided hit, only one direction matches
- `Driver` column tells you which channel is driving the score

## GSEA Validation

Validate hits by examining the running enrichment:

```{r gsea-validate, fig.cap="GSEA validation plots for a top hit showing Up and Down gene enrichment."}
# Get logFC for a top hit
candidate <- out_ud$res$signature_id[2]  # Second-ranked hit
gl_cand <- gpsa_get_signature(db, candidate, dataset = "logFC")

# Plot enrichment for Up genes
gpsa_plot_gsea(gl_cand, query_ud$Up$genes,
               main = paste0(candidate, " vs Query Up genes"))

# Plot enrichment for Down genes
gpsa_plot_gsea(gl_cand, query_ud$Down$genes,
               main = paste0(candidate, " vs Query Down genes"))
```

## Best Practices

1. **Start with fixed selection** (`n_each = 500`) for initial exploration
2. **Use adaptive selection** when you want interpretable, energy-based cutoffs
3. **Always examine channel scores** to understand what's driving the hit
4. **Use dual-mode scoring** when you need to distinguish phenocopy from discovery
5. **Validate top hits** with GSEA plots to confirm biological relevance

## Summary

| Approach | Function | When to Use |
|----------|----------|-------------|
| Fixed 500 | `gpsa_make_updown_query(gl, n_each = 500)` | Quick exploration, standard cMAP |
| Fixed custom | `gpsa_make_updown_query(gl, n_each = 300)` | Smaller/larger gene sets |
| Adaptive | `gpsa_make_updown_query_adaptive(gl, eta = 0.90)` | Interpretable cutoffs |

## Session Info

```{r session}
sessionInfo()
```
