---
title: "Tutorial: Full Vector Search"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Tutorial: Full Vector Search}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
# Resolve HDF5 database path (for vignette rendering)
resolve_h5_path <- function() {
  h5 <- Sys.getenv("GPSA_H5", unset = "")
  if (h5 != "" && file.exists(h5)) return(h5)
  
  h5 <- getOption("GPSA.h5", default = "")
  if (h5 != "" && file.exists(h5)) return(h5)
  
  candidates <- c(
    "../output/gpsa_rankZ_logFC_optimized.h5",
    "output/gpsa_rankZ_logFC_optimized.h5",
    "~/gpsa_db.h5"
  )
  for (h5 in candidates) {
    if (file.exists(h5)) return(normalizePath(h5))
  }
  return("")
}

h5_path <- resolve_h5_path()
has_db <- (h5_path != "")

knitr::opts_chunk$set(
  eval = has_db,
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)

if (has_db) {
  options(GPSA.h5 = h5_path)
}
```

## Overview

While Up/Down queries use only the most extreme genes, **vector search** uses the entire gene expression profile as a query. This approach:

- Leverages information from all genes (not just top/bottom)
- Uses rankZ transformation for robust similarity
- Supports adaptive sparsification to balance signal and noise

## Setup

```{r load}
library(GPSA)

# Load database
db <- gpsa_load_db(preload_Z = TRUE, verbose = TRUE)
```

## Preparing a Full Vector Query

### Extract Complete Expression Profile

```{r get-vector}
# Get a reference signature's full logFC vector
sig_ref <- db$sigs[1]  # Use first signature as example
gl_full <- gpsa_get_signature(db, sig_ref, dataset = "logFC")

# This contains logFC values for all genes
message(sprintf("Vector length: %d genes", length(gl_full)))
print(summary(gl_full))
```

### Understanding the Query Vector

```{r examine-vector, fig.cap="Distribution of logFC values in the reference signature."}
# The vector is named by gene symbol
message("First 6 genes:")
print(head(gl_full))

# Distribution of logFC values
hist(gl_full, breaks = 50, main = "logFC Distribution",
     xlab = "log2 Fold Change", col = "steelblue")

# How many genes have substantial changes?
message(sprintf("Genes with |logFC| > 1: %d", sum(abs(gl_full) > 1)))
message(sprintf("Genes with |logFC| > 0.5: %d", sum(abs(gl_full) > 0.5)))
```

## Running Vector Search

### Basic Vector Search

```{r vector-search}
out_vec <- gpsa_search_vector(
  db, 
  gl_full,
  eta = 0.90,           # Capture 90% of signal energy
  min_k = 300,          # Minimum 300 genes
  max_k = 5000,         # Maximum 5000 genes
  transform = "rankZ",  # Apply rankZ transformation
  normalize = TRUE      # Normalize to unit length
)

# Check self-rank (should be very high)
self_rank <- gpsa_rank_in(out_vec$res, sig_ref)
message(sprintf("Self-rank in vector search: %d", self_rank))
```

### Examining Compile Information

```{r compile-info}
# The compile info shows how the vector was processed
message("=== Vector Search Compile Info ===")
print(out_vec$compile)

# Key fields:
# - k_used: number of genes retained after sparsification
# - energy_captured: fraction of signal energy retained
# - threshold: logFC threshold for sparsification
```

### View Results

```{r view-results}
message("=== Vector Search Top 10 ===")
print(head(out_vec$res[, c("signature_id", "pbgene", "CellLineName",
                           "Score", "tau", "FDR_rank")], 10))
```

## Understanding the Parameters

### Energy Threshold (eta)

The `eta` parameter controls how much of the signal to retain:

```{r eta-examples}
# High eta (0.95): Keep more genes, capture more signal
out_95 <- gpsa_search_vector(db, gl_full, eta = 0.95, min_k = 300)
message(sprintf("eta=0.95: %d genes used", out_95$compile$k_used))

# Lower eta (0.80): Keep fewer genes, focus on strongest signals
out_80 <- gpsa_search_vector(db, gl_full, eta = 0.80, min_k = 300)
message(sprintf("eta=0.80: %d genes used", out_80$compile$k_used))
```

### Min/Max Gene Constraints

```{r min-max}
# Force at least 500 genes
out_min500 <- gpsa_search_vector(db, gl_full, eta = 0.90, min_k = 500)
message(sprintf("min_k=500: %d genes used", out_min500$compile$k_used))

# Limit to at most 1000 genes
out_max1000 <- gpsa_search_vector(db, gl_full, eta = 0.90, max_k = 1000)
message(sprintf("max_k=1000: %d genes used", out_max1000$compile$k_used))
```

### Transform Options

```{r transform, eval=FALSE}
# rankZ (default): Robust to outliers, cMAP-style semantics
out_rankz <- gpsa_search_vector(db, gl_full, transform = "rankZ")

# none: Use raw values (if already transformed)
# out_raw <- gpsa_search_vector(db, gl_full, transform = "none")
```

## Comparing Vector Search vs Up/Down Query

```{r compare-methods}
# Up/Down query with same reference
query_ud <- gpsa_make_updown_query(gl_full, n_each = 500)
out_ud <- gpsa_search(db, query_ud, return_components = FALSE)

# Compare self-ranks
self_vec <- gpsa_rank_in(out_vec$res, sig_ref)
self_ud <- gpsa_rank_in(out_ud$res, sig_ref)

message("Self-rank comparison:")
message(sprintf("  Vector search: %d", self_vec))
message(sprintf("  Up/Down query: %d", self_ud))

# Compare top hits overlap
top_vec <- head(out_vec$res$signature_id, 100)
top_ud <- head(out_ud$res$signature_id, 100)
overlap <- length(intersect(top_vec, top_ud))

message(sprintf("Top 100 overlap: %d signatures", overlap))
```

### When to Use Each Approach

| Approach | Best For |
|----------|----------|
| Up/Down Query | Classic cMAP analysis, interpretable results |
| Vector Search | Maximum information usage, subtle signatures |
| Adaptive Up/Down | Balance between both approaches |

## Adaptive Sparsification Deep Dive

### How It Works

1. Compute squared values (energy) for each gene
2. Sort genes by energy (descending)
3. Cumulative sum until reaching eta fraction of total energy
4. Retain those genes, zero out the rest

```{r sparsification-viz}
# The compile info from gpsa_search_vector shows sparsification details
message("=== Sparsification Summary ===")
message(sprintf("Genes used: %d", out_vec$compile$k))
message(sprintf("Energy captured: %.1f%%", out_vec$compile$capture * 100))
```

### Examining Retained Genes

```{r retained-genes}
# Top genes by absolute logFC (these are most likely retained)
top_genes <- sort(abs(gl_full), decreasing = TRUE)[1:20]
message("=== Top 20 Genes by |logFC| ===")
print(top_genes)
```

## Using Custom Gene Lists

You can also use your own gene list (not from database):

```{r custom-genelist, eval=FALSE}
# Create a custom geneList from your experiment
# Named vector: names = gene symbols, values = logFC
my_genelist <- c(
  "TP53" = 2.5,
  "BRCA1" = 1.8,
  "EGFR" = -1.2,
  "MYC" = -2.0
  # ... more genes
)

# Search with custom list
# Note: genes not in database will be ignored
out_custom <- gpsa_search_vector(
  db, 
  my_genelist,
  eta = 0.90,
  min_k = 10,  # Lower min for small gene lists
  transform = "rankZ"
)
```

## Handling Large Gene Lists

For very large input vectors:

```{r large-vectors}
# If you have expression for all genes
# The function automatically:
# 1. Filters to genes present in database
# 2. Applies sparsification
# 3. Normalizes the query vector

# Check overlap with database genes
overlap_genes <- intersect(names(gl_full), db$genes)
message(sprintf("Genes in common with database: %d", length(overlap_genes)))
```

## Best Practices

1. **Start with default parameters**: `eta = 0.90`, `min_k = 300`, `max_k = 5000`
2. **Compare with Up/Down**: Vector search should give similar but potentially better results
3. **Check compile info**: Understand how many genes are actually used
4. **Use higher eta for subtle signatures**: If changes are small, retain more genes
5. **Validate top hits**: Always examine GSEA plots for biological relevance

## Performance Notes

```{r performance, eval=FALSE}
# Vector search is slightly slower than Up/Down due to:
# 1. Energy computation for all genes
# 2. Larger query vectors (more genes)
# But still very fast in memory mode (~0.2s)
```

## Summary

| Parameter | Default | Purpose |
|-----------|---------|---------|
| `eta` | 0.90 | Fraction of signal energy to capture |
| `min_k` | 300 | Minimum genes to retain |
| `max_k` | 5000 | Maximum genes to retain |
| `transform` | "rankZ" | How to transform query values |
| `normalize` | TRUE | Normalize to unit length |

## Session Info

```{r session}
sessionInfo()
```
